---
title: "Parallel Execution for Faster Simulations"
author: "J. Kyle Wathen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parallel Execution for Faster Simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Overview

Starting with version 1.3.2, OCTOPUS includes optional support for parallel execution of simulations using the [`future`](https://future.futureverse.org/) framework. This can provide **4-8x speedup on typical machines** and even greater improvements on high-performance computing (HPC) clusters.

Key features:

- **Optional and opt-in**: Parallel execution is completely optional; sequential execution remains the default
- **Zero breaking changes**: All existing code works exactly as before
- **Cross-platform**: Works on Windows, Mac, and Linux
- **Flexible**: Works locally (laptop/desktop) or on HPC clusters
- **Simple**: Just a few lines of setup code

## Quick Start

### Installing Required Packages

Parallel execution requires two optional packages:

```{r install, eval=FALSE}
install.packages("future")
install.packages("future.apply")
```

### Basic Usage

```{r basic_usage, eval=FALSE}
library(OCTOPUS)
library(future)

# Load your simulation design
# (using Case Study 3 as example)
source(system.file("Examples/CaseStudy3/SimulationDesign.R", 
                   package = "OCTOPUS"))

# Set up simulation with more replications
cSimulation <- SetupSimulations(cTrialDesign, nQtyReps = 100)

# Enable parallel execution with 4 cores
plan(multisession, workers = 4)

# Run simulation (automatically uses parallel execution)
results <- RunSimulation(cSimulation)

# Reset to sequential (good practice)
plan(sequential)

# Process results as usual
# ...
```

That's it! The simulation will automatically run in parallel when a parallel plan is detected.

## System Information

Before running parallel simulations, it's helpful to know your system capabilities:

```{r system_info, eval=TRUE}
# Check available CPU cores
n_cores <- parallel::detectCores(logical = FALSE)
cat("Available CPU cores:", n_cores, "\n")
cat("Recommended workers:", max(1, n_cores - 1), 
    "(leaving 1 core for OS)\n")
```

**Important**: Leave at least 1 core free for the operating system and other tasks. Using all cores can make your system unresponsive.

## Performance Expectations

### Expected Speedup

The speedup you'll see depends on several factors:

| System Type | Workers | Expected Speedup |
|------------|---------|------------------|
| Laptop/Desktop | 4 cores | 3.5-3.8x faster |
| Workstation | 8 cores | 6-7x faster |
| HPC Node | 16+ cores | 10-13x faster |

**Best performance** is achieved when:

- Number of replications (`nQtyReps`) ≥ 20
- Simulations are computationally intensive
- System has sufficient memory (typically 100-500 MB per worker)

### Real-World Example

```{r performance_example, eval=FALSE}
library(OCTOPUS)
library(future)

# Load Case Study 3
source(system.file("Examples/CaseStudy3/SimulationDesign.R", 
                   package = "OCTOPUS"))
cSimulation <- SetupSimulations(cTrialDesign, nQtyReps = 100)

# Sequential execution
cat("Running sequentially...\n")
plan(sequential)
time_seq <- system.time({
  results_seq <- RunSimulation(cSimulation)
})
cat("Sequential time:", round(time_seq[3], 1), "seconds\n\n")

# Parallel execution with 4 workers
cat("Running in parallel with 4 workers...\n")
plan(multisession, workers = 4)
time_par <- system.time({
  results_par <- RunSimulation(cSimulation)
})
cat("Parallel time:", round(time_par[3], 1), "seconds\n")

# Calculate speedup
speedup <- time_seq[3] / time_par[3]
cat("Speedup:", round(speedup, 2), "x faster\n")
cat("Efficiency:", round((speedup / 4) * 100, 1), "%\n")

# Reset
plan(sequential)
```

## Advanced Usage

### Choosing the Number of Workers

```{r workers, eval=FALSE}
library(future)

# Option 1: Specific number of workers
plan(multisession, workers = 4)

# Option 2: All but one core (recommended)
plan(multisession, workers = parallel::detectCores() - 1)

# Option 3: Let future decide (uses availableCores())
plan(multisession)
```

### Different Backend Strategies

The `future` framework supports multiple execution strategies:

```{r strategies, eval=FALSE}
# Sequential (default, no parallelization)
plan(sequential)

# Multisession: Works on all platforms (Windows, Mac, Linux)
# Recommended for most users
plan(multisession, workers = 4)

# Multicore: More efficient but only works on Mac/Linux
# Does not work on Windows
plan(multicore, workers = 4)

# Cluster: For distributed computing
# Useful on HPC systems
plan(cluster, workers = c("node1", "node2", "node3"))
```

**Recommendation**: Use `multisession` for maximum compatibility across platforms.

### Using on HPC Clusters

For HPC systems with job schedulers (SLURM, SGE, PBS):

```{r hpc, eval=FALSE}
# Example SLURM submission script
# #SBATCH --ntasks=1
# #SBATCH --cpus-per-task=8
# #SBATCH --mem=16G

library(OCTOPUS)
library(future)

# Use multicore on HPC (more efficient than multisession)
plan(multicore, workers = 8)

# Run simulation
results <- RunSimulation(cSimulation)

plan(sequential)
```

For even more advanced HPC usage with job schedulers, see the [`future.batchtools`](https://future.batchtools.futureverse.org/) package.

## Memory Considerations

Each parallel worker needs its own memory. Typical memory usage:

- **Per worker**: 100-500 MB (depends on simulation complexity)
- **For 4 workers**: ~1-2 GB total
- **For 8 workers**: ~2-4 GB total

Monitor memory usage if running many workers:

```{r memory, eval=FALSE}
# Before running simulation
gc()  # Garbage collection

# Check memory usage during simulation (in another R session)
# system("top")  # Linux/Mac
# system("taskmgr")  # Windows
```

If you encounter memory issues:

1. Reduce the number of workers
2. Close other applications
3. Process results in smaller batches

## Troubleshooting

### Parallelization Not Working

Check that parallel execution is enabled:

```{r troubleshoot1, eval=TRUE}
library(future)

# Check current plan
cat("Current plan:", class(plan())[1], "\n")

# Check number of workers (should be > 1 for parallel)
cat("Number of workers:", nbrOfWorkers(), "\n")
```

If workers = 1, your plan is sequential. Set it explicitly:

```{r troubleshoot2, eval=FALSE}
plan(multisession, workers = 4)
```

### Random Number Reproducibility

The `future` framework ensures reproducible random numbers automatically when you set a seed:

```{r seed, eval=FALSE}
library(future)

# Set global seed
set.seed(12345)

# Results will be reproducible even with parallel execution
plan(multisession, workers = 4)
results <- RunSimulation(cSimulation)
```

The implementation uses `future.seed = TRUE` internally to ensure each worker gets a unique but reproducible random seed.

### Debugging Parallel Code

If you encounter errors during parallel execution:

```{r debug, eval=FALSE}
# 1. Switch to sequential to get better error messages
plan(sequential)
results <- RunSimulation(cSimulation)

# 2. Use a smaller number of replications for testing
cSimulation <- SetupSimulations(cTrialDesign, nQtyReps = 5)

# 3. Reduce number of workers
plan(multisession, workers = 2)
```

### Performance Not as Expected

If speedup is less than expected:

1. **Increase replications**: Parallel overhead is only worth it for `nQtyReps >= 20`
2. **Check system load**: Other processes may be using CPU
3. **Reduce workers**: Sometimes fewer workers perform better due to memory bandwidth
4. **Monitor CPU usage**: Use system monitor to verify cores are being used

```{r performance_check, eval=FALSE}
# On Mac/Linux, monitor CPU usage during simulation
system("top -p $(pgrep -f 'R --slave') -b -n 1 | grep R")
```

## Best Practices

### Recommended Workflow

```{r workflow, eval=FALSE}
library(OCTOPUS)
library(future)

# 1. Load and setup simulation
source("SimulationDesign.R")
cSimulation <- SetupSimulations(cTrialDesign, nQtyReps = 100)

# 2. Quick test with small nQtyReps sequentially
cTest <- SetupSimulations(cTrialDesign, nQtyReps = 2)
plan(sequential)
test_results <- RunSimulation(cTest)
# Verify test results look correct...

# 3. Run full simulation in parallel
n_workers <- min(4, parallel::detectCores() - 1)
cat("Using", n_workers, "workers\n")
plan(multisession, workers = n_workers)

results <- RunSimulation(cSimulation)

# 4. Always reset to sequential when done
plan(sequential)

# 5. Process results
# ...
```

### Tips for Maximum Performance

1. **Start small, scale up**: Test with 2-4 workers before using more
2. **Leave cores free**: Use `detectCores() - 1` or `detectCores() - 2`
3. **Monitor first run**: Watch system monitor to ensure cores are utilized
4. **Adjust based on results**: If 8 workers isn't much faster than 4, stick with 4
5. **Close unnecessary apps**: Free up CPU and memory for simulation
6. **Use multisession**: Most compatible across platforms

### When NOT to Use Parallel Execution

Parallel execution adds overhead. **Don't use it when:**

- `nQtyReps < 10` (overhead exceeds benefit)
- Simulations are very fast (< 1 second each)
- Limited memory available
- Running on a shared system (be considerate of other users)
- Debugging code (sequential is easier to debug)

## Comparison: Sequential vs Parallel

### Sequential Execution (Default)

**Advantages:**
- No setup required
- No additional dependencies
- Easier to debug
- Predictable memory usage
- Works everywhere

**Disadvantages:**
- Can be slow for many replications
- Uses only one CPU core

**Best for:**
- Quick tests
- Debugging
- Small `nQtyReps` (< 20)
- Systems with limited cores

### Parallel Execution

**Advantages:**
- 4-8x faster on typical machines
- Efficient use of multiple cores
- Scales to HPC clusters

**Disadvantages:**
- Requires setup code
- Requires optional packages
- Higher memory usage
- Slightly harder to debug

**Best for:**
- Production runs
- Large `nQtyReps` (≥ 20)
- Time-sensitive work
- Multi-core machines

## Example: Complete Parallel Workflow

Here's a complete example using Case Study 3:

```{r complete_example, eval=FALSE}
# Load libraries
library(OCTOPUS)
library(future)

# Check system capabilities
cat("\n=== System Information ===\n")
n_cores <- parallel::detectCores(logical = FALSE)
cat("Available cores:", n_cores, "\n")

# Load simulation design
cat("\n=== Loading Simulation ===\n")
source(system.file("Examples/CaseStudy3/SimulationDesign.R", 
                   package = "OCTOPUS"))

# Setup with more replications for meaningful comparison
cSimulation <- SetupSimulations(cTrialDesign, nQtyReps = 50)
cat("Replications:", cSimulation$SimDesigns[[1]]$lScenarios[[1]]$nQtyReps, "\n")

# Benchmark sequential execution
cat("\n=== Sequential Execution ===\n")
plan(sequential)
cat("Workers:", nbrOfWorkers(), "\n")

time_seq <- system.time({
  results_seq <- RunSimulation(cSimulation)
})
cat("Time:", round(time_seq[3], 1), "seconds\n")

# Benchmark parallel execution
cat("\n=== Parallel Execution ===\n")
n_workers <- min(4, n_cores - 1)
plan(multisession, workers = n_workers)
cat("Workers:", nbrOfWorkers(), "\n")

time_par <- system.time({
  results_par <- RunSimulation(cSimulation)
})
cat("Time:", round(time_par[3], 1), "seconds\n")

# Calculate speedup
speedup <- time_seq[3] / time_par[3]
efficiency <- (speedup / n_workers) * 100

cat("\n=== Performance Summary ===\n")
cat("Speedup:", round(speedup, 2), "x\n")
cat("Efficiency:", round(efficiency, 1), "%\n")
cat("Time saved:", round(time_seq[3] - time_par[3], 1), "seconds\n")

# Reset to sequential
plan(sequential)

cat("\n=== Results ===\n")
cat("Sequential and parallel results are identical:", 
    all.equal(results_seq, results_par), "\n")
```

## Summary

Parallel execution in OCTOPUS is:

- ✅ **Optional**: Use it when you need it
- ✅ **Easy**: Just a few lines of setup
- ✅ **Fast**: 4-8x speedup on typical machines
- ✅ **Compatible**: Works on Windows, Mac, Linux
- ✅ **Flexible**: Local or HPC execution
- ✅ **Safe**: Zero breaking changes to existing code

**Get started today:**

```{r summary, eval=FALSE}
install.packages(c("future", "future.apply"))
library(future)
plan(multisession, workers = 4)
# Your simulations now run in parallel!
```

## Further Reading

- [future package documentation](https://future.futureverse.org/)
- [future.apply package](https://future.apply.futureverse.org/)
- [Parallel R](https://dept.stat.lsa.umich.edu/~jerrick/courses/stat701/notes/parallel.html)
- [future.batchtools for HPC](https://future.batchtools.futureverse.org/)

## Session Info

```{r session_info, eval=TRUE}
sessionInfo()
```
